---
title: "Abortion Analysis"
output: html_document
date: '2022-05-04'
by: Ryan Little
---

##### Packages and Functions

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(janitor)
library(tools)
options(digits=3)
options(scipen=999)
'%notin%' <- Negate('%in%')

#does a standard group_by and count() with percentage
grouper <- function(input_df, group_by_column, new_column_name = "n()"){
  output_df <- input_df %>%
    group_by(.data[[group_by_column]]) %>%
    summarise(temp_count = n()) %>%
    mutate(percent = temp_count/sum(temp_count)*100) %>%
    arrange(desc(percent)) %>%
    rename(!!new_column_name := temp_count)
  return(output_df)
}

#group/counts every column in input dataframe
group_count <- function(input_df, group_column_name='n()', state_filter=NA, start_col = 1){
  column_names <- colnames(input_df)
  if(!is.na(state_filter)){
    input_df <- input_df %>%
      filter(state == state_filter)
  }
  for (column in column_names[start_col:length(column_names)]){
    output <- grouper(input_df, column, group_column_name)
    print(output)
  }
}

grouper_sum <- function(input_df, group_by_column, sum_column, new_column_name = "n()"){
  output_df <- input_df %>%
    group_by(.data[[group_by_column]]) %>%
    summarise(temp_count = sum(.data[[sum_column]])) %>%
    mutate(percent = temp_count/sum(temp_count)*100) %>%
    arrange(desc(percent)) %>%
    rename(!!new_column_name := temp_count)
  return(output_df)
  
}

```

##### Import Data

```{r, message = FALSE, warning = FALSE}

#data set taken from https://www.cdc.gov/mmwr/volumes/70/ss/ss7009a1.htm#T1_down
import <- read_csv("data/cdc-reproductive-health-by-state.csv") %>%
  clean_names() %>%
  #cleaning random characters out
  mutate(state_3 = gsub("[^[:alnum:]]", " ", state_3),
         state_8 = gsub("[^[:alnum:]]", " ", state_8),
         state_13 = gsub("[^[:alnum:]]", " ", state_13),
         state_16 = gsub("[^[:alnum:]]", " ", state_16),
         state_22 = gsub("[^[:alnum:]]", " ", state_22),
         state_27 = gsub("[^[:alnum:]]", " ", state_27),
         state_35 = gsub("[^[:alnum:]]", " ", state_35)) %>%
  select(-state_3, -state_8, -state_13, -state_16, -state_22, -state_27, -state_35) %>%
  #separating counts and perc
  separate(perc_obtain_by_out_of_state, c("perc_obtain_by_out_of_state", "perc_obtain_by_out_of_state_perc"), " \\(") %>%
  separate(white, c("white", "white_perc"), " \\(") %>%
  separate(black, c("black", "black_perc"), " \\(") %>%
  separate(other, c("other", "other_perc"), " \\(") %>%
  separate(hispanic, c("hispanic", "hispanic_perc"), " \\(") %>%
  separate(married, c("married", "married_perc"), " \\(") %>%
  separate(unmarried, c("unmarried", "unmarried_perc"), " \\(") %>%
  separate(prev_births_0, c("prev_births_0", "prev_births_0_perc"), " \\(") %>%
  separate(prev_births_1, c("prev_births_1", "prev_births_1_perc"), " \\(") %>%
  separate(prev_births_2, c("prev_births_2", "prev_births_2_perc"), " \\(") %>%
  separate(prev_births_3, c("prev_births_3", "prev_births_3_perc"), " \\(") %>%
  separate(prev_births_4_or_more, c("prev_births_4_or_more", "prev_births_4_or_more_perc"), " \\(") %>%
  separate(prev_abortions_0, c("prev_abortions_0", "prev_abortions_0_perc"), " \\(") %>%
  separate(prev_abortions_1, c("prev_abortions_1", "prev_abortions_1_perc"), " \\(") %>%
  separate(prev_abortions_2, c("prev_abortions_2", "prev_abortions_2_perc"), " \\(") %>%
  separate(prev_abortions_3_or_more, c("prev_abortions_3_or_more", "prev_abortions_3_or_more_perc"), " \\(") %>%
  separate(weeks_gestation_less_6, c("weeks_gestation_less_6", "weeks_gestation_less_6_perc"), " \\(") %>%
  separate(weeks_gestation_7_to_9, c("weeks_gestation_7_to_9", "weeks_gestation_7_to_9_perc"), " \\(") %>%
  separate(weeks_gestation_10_13, c("weeks_gestation_10_13", "weeks_gestation_10_13_perc"), " \\(") %>%
  separate(weeks_gestation_14_15, c("weeks_gestation_14_15", "weeks_gestation_14_15_perc"), " \\(") %>%
  separate(weeks_gestation_16_17, c("weeks_gestation_16_17", "weeks_gestation_16_17_perc"), " \\(") %>%
  separate(weeks_gestation_18_20, c("weeks_gestation_18_20", "weeks_gestation_18_20_perc"), " \\(") %>%
  separate(weeks_gestation_21_or_more, c("weeks_gestation_21_or_more", "weeks_gestation_21_or_more_perc"), " \\(") %>%
  separate(surgical_less_than_13_weeks, c("surgical_less_than_13_weeks", "surgical_less_than_13_weeks_perc"), " \\(") %>%
  separate(surgical_more_than_13_weeks, c("surgical_more_than_13_weeks", "surgical_more_than_13_weeks_perc"), " \\(") %>%
  separate(surgical_unknown, c("surgical_unknown", "surgical_unknown_perc"), " \\(") %>%
  separate(medical_less_than_9_weeks, c("medical_less_than_9_weeks", "medical_less_than_9_weeks_perc"), " \\(") %>%
  separate(medical_more_than_9_weeks, c("medical_more_than_9_weeks", "medical_more_than_9_weeks_perc"), " \\(") %>%
  separate(medical_unknown, c("medical_unknown", "medical_unknown_perc"), " \\(") %>%
  separate(intrauterin_instillation, c("intrauterin_instillation", "intrauterin_instillation_perc"), " \\(") %>%
  separate(hysterectomy_hysterotomy, c("hysterectomy_hysterotomy", "hysterectomy_hysterotomy_perc"), " \\(") %>%
  #removing ) and making double
  mutate(perc_obtain_by_out_of_state_perc = as.double(gsub("\\)", "", perc_obtain_by_out_of_state_perc)),
           white_perc = as.double(gsub("\\)", "", white_perc)),
           black_perc = as.double(gsub("\\)", "", black_perc)),
           other_perc = as.double(gsub("\\)", "", other_perc)),
           hispanic_perc = as.double(gsub("\\)", "", hispanic_perc)),
           married_perc = as.double(gsub("\\)", "", married_perc)),
           unmarried_perc = as.double(gsub("\\)", "", unmarried_perc)),
           prev_births_0_perc = as.double(gsub("\\)", "", prev_births_0_perc)),
           prev_births_1_perc = as.double(gsub("\\)", "", prev_births_1_perc)),
           prev_births_2_perc = as.double(gsub("\\)", "", prev_births_2_perc)),
           prev_births_3_perc = as.double(gsub("\\)", "", prev_births_3_perc)),
           prev_births_4_or_more_perc = as.double(gsub("\\)", "", prev_births_4_or_more_perc)),
           prev_abortions_0_perc = as.double(gsub("\\)", "", prev_abortions_0_perc)),
           prev_abortions_1_perc = as.double(gsub("\\)", "", prev_abortions_1_perc)),
           prev_abortions_2_perc = as.double(gsub("\\)", "", prev_abortions_2_perc)),
           prev_abortions_3_or_more_perc = as.double(gsub("\\)", "", prev_abortions_3_or_more_perc)),
           weeks_gestation_less_6_perc = as.double(gsub("\\)", "", weeks_gestation_less_6_perc)),
           weeks_gestation_7_to_9_perc = as.double(gsub("\\)", "", weeks_gestation_7_to_9_perc)),
           weeks_gestation_10_13_perc = as.double(gsub("\\)", "", weeks_gestation_10_13_perc)),
           weeks_gestation_14_15_perc = as.double(gsub("\\)", "", weeks_gestation_14_15_perc)),
           weeks_gestation_16_17_perc = as.double(gsub("\\)", "", weeks_gestation_16_17_perc)),
           weeks_gestation_18_20_perc = as.double(gsub("\\)", "", weeks_gestation_18_20_perc)),
           weeks_gestation_21_or_more_perc = as.double(gsub("\\)", "", weeks_gestation_21_or_more_perc)),
           surgical_less_than_13_weeks_perc = as.double(gsub("\\)", "", surgical_less_than_13_weeks_perc)),
           surgical_more_than_13_weeks_perc = as.double(gsub("\\)", "", surgical_more_than_13_weeks_perc)),
           surgical_unknown_perc = as.double(gsub("\\)", "", surgical_unknown_perc)),
           medical_less_than_9_weeks_perc = as.double(gsub("\\)", "", medical_less_than_9_weeks_perc)),
           medical_more_than_9_weeks_perc = as.double(gsub("\\)", "", medical_more_than_9_weeks_perc)),
           medical_unknown_perc = as.double(gsub("\\)", "", medical_unknown_perc)),
           intrauterin_instillation_perc = as.double(gsub("\\)", "", intrauterin_instillation_perc)),
           hysterectomy_hysterotomy_perc = as.double(gsub("\\)", "", hysterectomy_hysterotomy_perc))
           ) %>%
  #making double
  mutate(perc_obtain_by_out_of_state = as.double(perc_obtain_by_out_of_state),
         white = as.double(white),
         black = as.double(black),
         other = as.double(other),
         hispanic  = as.double(hispanic),
         married = as.double(married),
         unmarried = as.double(unmarried),
         prev_births_0 = as.double(prev_births_0),
         prev_births_1 = as.double(prev_births_1),
         prev_births_2 = as.double(prev_births_2),
         prev_births_3 = as.double(prev_births_3),
         prev_births_4_or_more = as.double(prev_births_4_or_more),
         prev_abortions_0  = as.double(prev_abortions_0),
         prev_abortions_1  = as.double(prev_abortions_1),
         prev_abortions_2  = as.double(prev_abortions_2),
         prev_abortions_3_or_more  = as.double(prev_abortions_3_or_more),
         weeks_gestation_less_6  = as.double(weeks_gestation_less_6),
         weeks_gestation_7_to_9  = as.double(weeks_gestation_7_to_9),
         weeks_gestation_10_13 = as.double(weeks_gestation_10_13),
         weeks_gestation_14_15 = as.double(weeks_gestation_14_15),
         weeks_gestation_16_17 = as.double(weeks_gestation_16_17),
         weeks_gestation_18_20 = as.double(weeks_gestation_18_20),
         weeks_gestation_21_or_more  = as.double(weeks_gestation_21_or_more),
         surgical_less_than_13_weeks = as.double(surgical_less_than_13_weeks),
         surgical_more_than_13_weeks = as.double(surgical_more_than_13_weeks),
         surgical_unknown  = as.double(surgical_unknown),
         medical_less_than_9_weeks = as.double(medical_less_than_9_weeks),
         medical_more_than_9_weeks = as.double(medical_more_than_9_weeks),
         medical_unknown = as.double(medical_unknown),
         intrauterin_instillation  = as.double(intrauterin_instillation),
         hysterectomy_hysterotomy  = as.double(hysterectomy_hysterotomy)) %>%
  #changing to Title Case for viz
  mutate(cpr_rating = toTitleCase(cpr_rating))

#data from BigLocalNews/AP partnership
#importing census
census <- read_csv("data/01_state_pl94171_standard_compare_2010_2020.csv") %>%
  clean_names() %>%
  #selecting variables to analyze
  select(state_name, x2020_pop, x2020_pop_hisp, x2020_pop_white_non_hisp, x2020_pop_black_non_hisp, x2020_pop_aian_non_hisp, x2020_pop_nhpi_non_hisp, x2020_pop_other_non_hisp, x2020_land_area_sqmi, x2020_gq_inst_correctional, x2020_diversity_index, prop_change_pop, prop_change_pop_hisp, prop_change_pop_white_non_hisp, prop_change_pop_black_non_hisp, prop_change_pop_aian_non_hisp, prop_change_pop_asian_non_hisp, prop_change_pop_nhpi_non_hisp, prop_change_pop_other_non_hisp, prop_change_pop_two_plus_non_hisp)



#cleaning and joining
clean <- import %>%
  #joining census
  left_join(census, by = c("state_1" = "state_name")) %>%
  rename(state = state_1) %>%
  #calc pop in millions for viz
  mutate(pop_in_millions = x2020_pop/1000000) %>%
  #add all previous abortions to create one side of boolean
  mutate(previous_abortion_perc = prev_abortions_1_perc + prev_abortions_2_perc + prev_abortions_3_or_more_perc,
         previoss_birth_perc = prev_births_1_perc + prev_births_2_perc + prev_births_3_perc + prev_births_4_or_more_perc) %>%
  #pretty printing some stuff because flourish is a pain
  mutate(pretty_previous_abortion_perc = case_when(
          !is.na(previous_abortion_perc) ~ paste0(previous_abortion_perc, "%"),
          TRUE ~ 'No Data'),
         pretty_previoss_birth_perc = case_when(
          !is.na(previous_abortion_perc) ~ paste0(previoss_birth_perc, "%"),
          TRUE ~ 'No Data'),
         pretty_married_perc = case_when(
          !is.na(previous_abortion_perc) ~ paste0(married_perc, "%"),
          TRUE ~ 'No Data')
         ) %>%
  mutate(prisoners_per_100k = x2020_gq_inst_correctional/(x2020_pop/100000))

#writing output for viz
#write_csv(clean, "output/clean-abortion-data.csv")
```

# Factcheck

"The generally Southern and central states are home to 158 million Americans."

```{r}

grouper_sum(clean, "cpr_rating", "x2020_pop")

```


"The states encompass 47 percent of the United States land mass."

```{r}

grouper_sum(clean, "cpr_rating", "x2020_land_area_sqmi")

```

"A Center for Reproductive Rights analysis of laws state laws found that 25 states and three territories are “hostile” to abortion rights, meaning they could immediately prohibit abortion."

"Three more states plus Puerto Rico and the U.S. Virgin Islands have no existing state protections, making it unclear if legislatures there would enact a ban. Abortion rights are either protected by state law or guarantee expanded access in the 22 other states. "
  
```{r}

grouper(clean, "cpr_rating")

```

"These states also imprison people at rates nearly double those found in states that have expanded access to abortion, according to a Mother Jones analysis of 2020 Decennial Census."
 
```{r}

clean %>%
  group_by(cpr_rating) %>%
  summarise(prisoners_per_100k = mean(prisoners_per_100k)) %>%
  arrange(desc(prisoners_per_100k))

```

"Analysis excludes California, New Hampshire and Maryland because those states do not report data to the CDC."

```{r}

clean %>%
  group_by(state) %>%
  summarise(abortions = sum(abortions)) %>%
  filter(is.na(abortions))

```
  
"The 25 hostile states already had far fewer abortions per person than other states between 2010 and 2019, according to Centers for Disease Control and Prevention Abortion Surveillance data."
  
```{r}

clean %>%
  group_by(cpr_rating) %>%
  summarise(abortion_rates = mean(abortion_rate, na.rm=TRUE)) %>%
  arrange(desc(abortion_rates))

```
